---
title: "R package jmvReadWrite"
author: "Sebastian Jentschke"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{jmvReadWrite}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      comment=NA,message=FALSE,
                      warning=FALSE,
                      fig.height=5,
                      fig.width=7,
                      fig.align="center")
```

The R-package  `jmvReadWrite` reads and writes the .omv-files that are used by
the statistical spreadsheet `jamovi` (www.jamovi.org). It is supposed to ease
using syntax for statistical analyses created using the GUI in `jamovi` in
connection with the R-library `jmv`.


## Installation

You can install the development version of the `jmvReadWrite` package from
github.

```{r, eval=FALSE}
if(!require(devtools)) install.packages("devtools")
devtools::install_github("sjentsch/jmvReadWrite")
```

Alternatively, there is a stable version available on CRAN. It can be installed
using the following commands

```{r, eval=FALSE}
install.packages("jmvReadWrite")
```

 
## How to use the package?

The following code uses the ToothGrowth-data set that is part of the data sets
included in R (the current file contains some modifications though for testing
the reading and writing routines: `jmvRead` and `jmvWrite`). With this data
set, a syntax to conduct an ANOVA is run.

The results should be similar to those obtained when running the same analysis
in jamovi (using the GUI). To do so, open the file menu (☰) choose `Open`,
`Data Library` and `ToothGrowth`. Afterwards, click on the `ANOVA`-button in
the `Analyses`-tab and choose `ANOVA`. There, you assign the variable `len` to
`Dependent Variable` and `supp` and `dose` to `Fixed Factors`. Afterwards, you
choose / tick `Overall Model Test` and `ω²`. Open the drop-down menu
`Assumption Checks` and tick `Homogeneity test` and `Normality test`.
The results should be identical apart from that the table output looks nicer in
`jamovi` (not only text, as below), numbers are rounded and maybe one or two
other cosmetic differences.

If you want to copy the syntax generated in jamovi, you have to switch on the
[`Syntax Mode`](https://jamovi.readthedocs.io/en/latest/getting-started/um_2_first-steps.html#syntax-mode).
Afterwards, the syntax is shown at the top of the analysis and can be copied
from there.

```{r, echo=TRUE}
library(jmvReadWrite)
library(jmv)

data = jmvRead(fleNme = system.file("extdata", "ToothGrowth.omv", package = "jmvReadWrite"))
jmv::ANOVA(
    formula = len ~ supp + dose + supp:dose,
    data = data,
    effectSize = c("omega"),
    modelTest = TRUE,
    homo = TRUE,
    norm = TRUE)
```

The `jmvReadWrite`-package also enables you to write `.omv`-files in order to
use them in `jamovi`. Let's assume that you have a large collection of
log-files (e.g., from an experiment) that you compile and process (summarize,
filter, etc.) in R in order to later analyse them in `jamovi`. You will have
those processed log-files stored in a data frame (called, e.g., `data`) which
you then write to a file that you can open in jamovi afterwards. Although
jamovi reads R-data files (.RData, .rda, .rds) `jmvWrite` permits to store
`jamovi`-specific attributes (such as variable labels) in addition.

```{r, eval=FALSE}
library(jmvReadWrite)

jmvWrite(dtaFrm = data, fleNme = 'Trial.omv')
```
