---
title: "R package `jmvReadWrite`"
author: "Sebastian Jentschke"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{R package `jmvReadWrite`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  rmarkdown::html_vignette:
    keep_md: TRUE  
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      comment = "#>",
                      collapse = TRUE,
                      message = FALSE,                      
                      warning = FALSE,
                      fig.height = 5,
                      fig.width = 7,
                      fig.align = "center",
                      out.width = "100%")
```


The R-package `jmvReadWrite` reads and writes the .omv-files that are used by
the statistical spreadsheet `jamovi` (www.jamovi.org). It is supposed to ease
using syntax for statistical analyses created using the GUI in `jamovi` in
connection with the R-library `jmv`.


## Installation

You can either install a stable version of `jmvReadWrite` which is available on
[CRAN](https://cran.r-project.org/package=jmvReadWrite) using the following command:

```{r, eval=FALSE}
install.packages("jmvReadWrite")
```

or you can install the development version of the `jmvReadWrite` package from
[GitHub](https://github.com/sjentsch/jmvReadWrite):

```{r, eval=FALSE}
if(!require(devtools)) install.packages("devtools")
devtools::install_github("sjentsch/jmvReadWrite")
```

 
## How to use the package?

The following code uses the ToothGrowth-data set that is part of the data sets
included in R (the current file contains some modifications though for testing
the reading and writing routines: `read_omv` and `write_omv`). With this data
set, a syntax to conduct an ANOVA is run.

The results should be similar to those obtained when running the same analysis
in jamovi (using the GUI). To do so, open the file menu (☰) choose `Open`,
`Data Library` and `ToothGrowth`. Afterwards, click on the `ANOVA`-button in
the `Analyses`-tab and choose `ANOVA`. There, you assign the variable `len` to
`Dependent Variable` and `supp` and `dose` to `Fixed Factors`. Afterwards, you
choose / tick `Overall Model Test` and `ω²`. Open the drop-down menu
`Assumption Checks` and tick `Homogeneity test` and `Normality test`.
The results should be identical apart from that the table output looks nicer in
`jamovi` (not only text, as below), numbers are rounded and maybe one or two
other cosmetic differences.

If you want to copy the syntax generated in jamovi, you have to switch on the
[`Syntax Mode`](https://jamovi.readthedocs.io/en/latest/_pages/um_6_syntax_mode.html).
Afterwards, the syntax is shown at the top of the analysis and can be copied
from there.

```{r, echo=TRUE}
library(jmvReadWrite)
library(jmv)

data = read_omv(fleNme = system.file("extdata", "ToothGrowth.omv", package = "jmvReadWrite"))
jmv::ANOVA(
    formula = len ~ supp + dose + supp:dose,
    data = data,
    effectSize = c("omega"),
    modelTest = TRUE,
    homo = TRUE,
    norm = TRUE)
```

Since version 0.2.0, read_omv also extracts the syntax from analyses that you
may have conducted in the jamovi-GUI and that are stored in the .omv-file. To
extract them, you have to set the parameter `getSyn = TRUE` when calling
read_omv (default is `FALSE`). When the parameter is set, the analyses are
stored in the attribute `syntax`. They can be used as shown in the following
examples:

```{r, echo=TRUE}
library(jmvReadWrite)

data = read_omv(fleNme = system.file("extdata", "ToothGrowth.omv", package = "jmvReadWrite"), getSyn = TRUE)
# shows the syntax of the analyses from the .omv-file
attr(data, 'syntax')
# runs the command of the first analysis
eval(parse(text=attr(data, 'syntax')[[1]]))
# runs the command of the second analysis and assigns the output from that analysis to the variable result2
eval(parse(text=paste0('result2 = ', attr(data, 'syntax')[[2]])))
names(result2)
# → "main"      "assump"    "contrasts" "postHoc"   "emm" (the names of the five output tables)
```

The `jmvReadWrite`-package also enables you to write `.omv`-files in order to
use them in `jamovi`. Let's assume that you have a large collection of
log-files (e.g., from an experiment) that you compile and process (summarize,
filter, etc.) in R in order to later analyse them in `jamovi`. You will have
those processed log-files stored in a data frame (called, e.g., `data`) which
you then write to a file that you can open in jamovi afterwards. Although
jamovi reads R-data files (.RData, .rda, .rds) `write_omv` permits to store
`jamovi`-specific attributes (such as variable labels) in addition.

```{r, eval=FALSE}
library(jmvReadWrite)

write_omv(dtaFrm = data, fleNme = 'Trial.omv')
```
