---
title: "R package `jmvReadWrite`"
author: "Sebastian Jentschke"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{R package `jmvReadWrite`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  rmarkdown::html_vignette:
    keep_md: TRUE  
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      comment = "#>",
                      collapse = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      fig.height = 5,
                      fig.width = 7,
                      fig.align = "center",
                      out.width = "100%")
```


The R-package `jmvReadWrite` reads and writes the .omv-files that are used by
the statistical spreadsheet `jamovi` (<https://www.jamovi.org>). It is supposed
to ease using jamovi and R together, provide helper functions for some often 
required data management tasks, and to adjust and use syntax for statistical
analyses that were created using the GUI in `jamovi` in R (in connection with
the R-library `jmv`). More recently, `jmvReadWrite` became easily available
from within `jamovi` by becoming part of the `Rj` module (where you can use it
by writing R commands, documented below), and via the `jTransform` module that
provides a graphical user interface for most helper functions.

## Installation

In R, you can either install a stable version of `jmvReadWrite` which is
available on [CRAN](https://cran.r-project.org/package=jmvReadWrite) using the
following command:

```{r, eval=FALSE}
install.packages("jmvReadWrite")
```

or you can install the development version of the `jmvReadWrite` package from
[GitHub](https://github.com/sjentsch/jmvReadWrite/):

```{r, eval=FALSE}
if (!require(devtools)) install.packages("devtools")
devtools::install_github("sjentsch/jmvReadWrite")
```

 
## How to use the package?

[**read_omv**](https://sjentsch.github.io/jmvReadWrite/reference/read_omv.html)

The following code uses the ToothGrowth-data set that is part of the data sets
included in R (the current file contains some modifications though for testing
the reading and writing routines: `read_omv` and `write_omv`). With this data
set, a syntax to conduct an ANOVA is run.

The results should be similar to those obtained when running the same analysis
in jamovi (using the GUI). To do so, open the file menu (☰) choose `Open`,
`Data Library` and `ToothGrowth`. Afterwards, click on the `ANOVA`-button in
the `Analyses`-tab and choose `ANOVA`. There, you assign the variable `len` to
`Dependent Variable` and `supp` and `dose` to `Fixed Factors`. Afterwards, you
choose / tick `Overall Model Test` and `ω²`. Open the drop-down menu
`Assumption Checks` and tick `Homogeneity test` and `Normality test`.
The results should be identical apart from that the table output looks nicer in
`jamovi` (not only text, as below), numbers are rounded and maybe one or two
other cosmetic differences.

If you want to copy the syntax generated in jamovi, you have to switch on the
[`Syntax Mode`](https://jamovi.readthedocs.io/en/latest/usermanual/um_6_jamovi_and_R.html#syntax-mode).
Afterwards, the syntax is shown at the top of the analysis and can be copied
from there.

```{r, echo=TRUE}
library(jmvReadWrite)
library(jmv)

fleOMV <- system.file("extdata", "ToothGrowth.omv", package = "jmvReadWrite")
data <- read_omv(fleOMV)
# if the "jmv"-package is installed, we can run a test analysis with the data
if ("jmv" %in% rownames(installed.packages())) {
    jmv::ANOVA(
        formula = len ~ supp + dose + supp:dose,
        data = data,
        effectSize = c("omega"),
        modelTest = TRUE,
        homo = TRUE,
        norm = TRUE)
    }
```

Since version 0.2.0,
[`read_omv`](https://sjentsch.github.io/jmvReadWrite/reference/read_omv.html)
also extracts the syntax from analyses that you may have conducted in the
jamovi-GUI and that are stored in the .omv-file. To extract them, you have to
set the parameter `getSyn = TRUE` when calling
[`read_omv`](https://sjentsch.github.io/jmvReadWrite/reference/read_omv.html)
(default is `FALSE`). When the parameter is set, the analyses are stored in the
attribute `syntax`. They can be used as shown in the following examples:

```{r, echo=TRUE}
library(jmvReadWrite)
fleOMV <- system.file("extdata", "ToothGrowth.omv", package = "jmvReadWrite")
data <- read_omv(fleOMV, getSyn = TRUE)
# shows the syntax of the analyses from the .omv-file
# please note that syntax extraction may not work on all systems
# if the syntax couldn't be extracted, an empty list (length = 0) is returned,
# otherwise, the syntax of the analyses from the .omv-file is shown and
# the commands of the first and the second analysis are run, with the
# output of the second analysis assigned to the variable result2
if (length(attr(data, "syntax")) >= 2) {
    attr(data, "syntax")
    # if the "jmv"-package is installed, we can run the analyses in "syntax"
    if ("jmv" %in% rownames(installed.packages())) {
        eval(parse(text = attr(data, "syntax")[[1]]))
        eval(parse(text = paste0("result2 = ", attr(data, "syntax")[[2]])))
        names(result2)
        # -> "main"      "assump"    "contrasts" "postHoc"   "emm"
        # (the names of the five output tables)
    }
}
```
<br/>

[**write_omv**](https://sjentsch.github.io/jmvReadWrite/reference/write_omv.html)

The `jmvReadWrite`-package also enables you to write `.omv`-files in order to
use them in `jamovi`. Let's assume that you have a large collection of
log-files (e.g., from an experiment) that you compile and process (summarize,
filter, etc.) in R in order to later analyse them in `jamovi`. You will have
those processed log-files stored in a data frame (called, e.g., `data`) which
you then write to a file that you can open in jamovi afterwards.

```{r, echo=TRUE}
library(jmvReadWrite)

# use the data set "ToothGrowth" and, if it exists, write it as jamovi-file
# using write_omv()
data("ToothGrowth")
# "retDbg" has to be set in order to return debug information to wrtDta
wrtDta <- write_omv(ToothGrowth, "Trial.omv", retDbg = TRUE)
names(wrtDta)
# -> "mtaDta" "xtdDta" "dtaFrm"
# this debug information contains a list with the metadata ("mtaDta", e.g.,
# column and data type), the extended data ("xtdDta", e.g., variable lables),
# and the data frame (dtaFrm) for checking (understanding the file format) and
# debugging

# check whether the file was written to the disk, get the file information (size, etc.)
# and delete the file afterwards
list.files(".", "Trial.omv")
file.info("Trial.omv")
unlink("Trial.omv")
```

Although jamovi reads R-data files (.RData, .rda, .rds)
[`write_omv`](https://sjentsch.github.io/jmvReadWrite/reference/write_omv.html)
permits to store `jamovi`-specific attributes (such as variable labels) in
addition. Please note that if you are reading from an `.omv`-file in order to
write back to an `.omv`-file (perhaps after some modifications), it is
recommended to leave the `sveAtt`-attribute set to `TRUE` (which is the
default).

```{r, echo=TRUE}
# reading and writing a file with the "sveAtt"-parameter permits you to keep
# essential meta-data to ensure that the written file looks and works like the
# original file (plus you modifications)
fleOMV <- system.file("extdata", "ToothGrowth.omv", package = "jmvReadWrite")
data <- read_omv(fleOMV, sveAtt = TRUE)
# shows the names of the attributes for the whole data set (e.g., number of
# rows and columns) and the names of the attributes of the first column
names(attributes(data))
names(attributes(data[[1]]))
#
# perhaps do some modifications to the file here and write it back afterwards
write_omv(data, "Trial.omv")
unlink("Trial.omv")
```

If `Trial.omv` in the example above would have been kept, it should look like
the original file (plus your possible modifications). If you, e.g., added a new
column, you could adjust some attributes (e.g., to enforce a specific
`columnType` or `measurementType`): just look at how attributes are stored for
other columns.
<br /><br />

**Helper functions**

`jmvReadWrite` contains a number of helper functions that assist you with data
management tasks that are frequently required:

* [`arrange_cols_omv`](https://sjentsch.github.io/jmvReadWrite/reference/arrange_cols_omv.html):
  Re-arranges the columns of your data file in a requested order.

* [`convert_to_omv`](https://sjentsch.github.io/jmvReadWrite/reference/convert_to_omv.html):
  Converts data sets from other file formats into jamovi-format (this function
  may be helpful if you have to convert a larger amount of files).

* [`long2wide_omv`](https://sjentsch.github.io/jmvReadWrite/reference/long2wide_omv.html):
  Convert a data set from long to wide format (time points for repeated
  measurements are arranged as rows in the original and converted into
  columns).

* [`wide2long_omv`](https://sjentsch.github.io/jmvReadWrite/reference/wide2long_omv.html):
  Convert a data set from wide to long format (time points for repeated
  measurements are arranged as columns in the original and converted into
  rows).

* [`merge_cols_omv`](https://sjentsch.github.io/jmvReadWrite/reference/merge_cols_omv.html):
  Add variables from several data sets (i.e., the variables / columns in the
  second, etc. input data set are added as columns to the first data set).

* [`merge_rows_omv`](https://sjentsch.github.io/jmvReadWrite/reference/merge_rows_omv.html):
  Add cases from several data sets (i.e., the cases / rows in the second, etc.
  data set are added as rows to the first data set).

* [`sort_omv`](https://sjentsch.github.io/jmvReadWrite/reference/sort_omv.html):
  Sort a data set according to one or more variable(s).

* [`transpose_omv`](https://sjentsch.github.io/jmvReadWrite/reference/transpose_omv.html):
  Transpose a data set (make rows into columns and vice versa).
